### 2010-07-23 papers/icfp-2010-desc/paper_desc_levitation.tex
gaol: boiling 4.3 and 4.4

In this section, we will turn to a more ambitious generic operation on
datatypes. Given a functor, represented as a tagged description, we
build the free monad over this functor.

\newcommand{\FMFreeMonad}{\D{FreeMonad}}
\newcommand{\FMFreeMonadD}{\D{FreeMonadD}}
\newcommand{\FMVar}{\C{Var}}
\newcommand{\FMComposite}{\C{Composite}}

Let us recall the free monad construction in, say, Haskell. Given a
functor \texttt{f}, the free monad over \texttt{f} is defined by the
following datatype:
%
\begin{code}
data FreeMonad f x
    = Var x
    | Composite (f (FreeMonad f x))
\end{code}
%
Being an inductive type, this $\FMFreeMonad$ datatype is itself
defined by a pattern functor. It is given by:
%
\[
\FMFreeMonadD\: \V{F}\: \V{X}\: \V{Z} \mapsto \V{X} \mathop{\D{+}} \V{F} \V{Z}
\]

@@

In our setting, the free monad construction will take the functor as a
tagged description, a set $\V{X}$ of variables, and will compute the
tagged description of the corresponding free monad. Implementing this
function is surprisingly easy:
%
\[\stk{
\FreeMonad{\_} : \TagDesc \To \Set \To \TagDesc \\
\FreeMonad{\pair{\V{E}}{\V{D}}{}}\:\V{X} \mapsto
    \pair{\pair{\DVar{}}{\V{E}}{}}
         {\pair{\DSigma{\V{X}}{\DUnit}}{\V{D}}{}}{}
}\]
%
We simply add a constructor, $\SYMBDVar$, and define its argument to
be a $\DSigma{\V{X}}{\DUnit}$, that is an element of $\V{X}$. We keep
$\V{E}$ and $\V{D}$ as they were, hence leaving the other constructors
unchanged. Unfolding the interpretation of this definition, we
convince ourselves that this corresponds to the functor
$\FMFreeMonadD$. The fixpoint operation ties the knot and gives us
the full-blown free monad construction.

@@

Of course, we must equip the resulting datatypes with operations
delivering a monadic interface. As expected, \(\LAM{\x}\DVar{\x}\)
plays the r\^ole of \return, embedding variables into terms. The
\bind operation corresponds to \emph{substitution}. We will now
implement it, as a generic function.

@@

Our implementation will appeal to the $\cata$ function developed
previously. So, let us write down the types, and fill as many
arguments to $\cata$ as possible:
%
\[\stk{
\begin{array}{@{}ll}
\subst : & \PITEL{\V{D}}{\TagDesc}
           \PI{\V{X}, \V{Y}}{\Set} 
           (\V{X} \To \Mu{(\toDesc{\FreeMonad{\V{D}}{\V{Y}}})}) \To \\
         & \Mu{(\toDesc{\FreeMonad{\V{D}}{\V{X}}})} \To
           \Mu{(\toDesc{\FreeMonad{\V{D}}{\V{Y}}})} 
\end{array} \\
\subst\: \V{D}\: \V{X}\: \V{Y}\: \V{\sigma} \mapsto
  \cata\: (\toDesc{\FreeMonad{\V{D}}{\V{X}}})\: 
          \Mu{(\toDesc{\FreeMonad{\V{D}}{\V{Y}}})}\: 
          \SHED
}\]
%
We are left with implementing the algebra of the catamorphism. Its
role is to catch appearances of $\DVar{\V{x}}$ and replace them by
$\V{\sigma}\: \V{x}$. This corresponds to the following definition:
%
\[\stk{
\begin{array}{@{}ll}
\apply : & \PITEL{\V{D}}{\TagDesc} 
           \PI{\V{X}, \V{Y}}{\Set} 
           (\V{X} \To \Mu{(\toDesc{\FreeMonad{\V{D}}{\V{Y}}})}) \To \\
         & \descop{\toDesc{\FreeMonad{\V{D}}{\V{X}}}}{\Mu{(\toDesc{\FreeMonad{\V{D}}{\V{Y}}})}} \To
           \Mu{(\toDesc{\FreeMonad{\V{D}}{\V{Y}}})}
\\
\end{array} \\
\begin{array}{@{}l@{\:\mapsto\:\:}l}
\apply\: \V{D}\: \V{X}\: \V{Y}\: \V{\sigma}\: \pair{\SYMBDVar}{\V{x}}{}   & \V{\sigma}\: \V{x}                   \\
\apply\: \V{D}\: \V{X}\: \V{Y}\: \V{\sigma}\: \pair{\V{c}}{\V{xs}}{} & \Con{\pair{\V{c}}{\V{xs}}{}}
\end{array}
}\]

@@

Filling the sub-goal with $\apply\: \V{D}\: \V{X}\: \V{Y}\:
\V{\sigma}$ completes the implementation. To sum up, we have
implemented the free monad construction for an arbitrary tagged
description. This gives the developer the ability, for any datatype,
to extend it with a notion of variable. Then, we have equipped this
structure with the corresponding monadic operations, \bind and
\return. This construction is an example of type-indexed
datatype~\cite{hinze:generic-haskell}, as found in Generic Haskell:
from a datatype, we build a new datatype and equip it with its
structure.

>>>>

In this section, we try a more ambitious generic operation. Given a functor,
represented as a tagged description, we
build its free monad.
%
\newcommand{\FMFreeMonad}{\D{FreeMonad}}
\newcommand{\FMFreeMonadD}{\D{FreeMonadD}}
\newcommand{\FMVar}{\C{Var}}
\newcommand{\FMComposite}{\C{Composite}}
%
Let us recall this construction in, say, Haskell. Given a
functor \texttt{f}, the free monad over \texttt{f} is given:
%
\begin{code}
data FreeMonad f x
    = Var x
    | Composite (f (FreeMonad f x))
\end{code}
%
Provided \texttt{f} is an instance of \texttt{Functor}, we may
take \texttt{Var} for \texttt{return} and use \texttt{f}'s
\texttt{fmap} to define \texttt{>>=} as substitution.

Being an inductive type, $\FMFreeMonad$ arises
by a pattern functor:
%
\[
\FMFreeMonadD\: \V{F}\: \V{X}\: \V{Z} \mapsto \V{X} \mathop{\D{+}} \V{F}\:\V{Z}
\]

@@

Our free monad construction takes the functor as a
tagged description, and given a set $\V{X}$ of variables, computes the
tagged description of the free monad pattern functor.
%
\[\stk{
\FreeMonad{\_} : \TagDesc \To \Set \To \TagDesc \\
\FreeMonad{\pair{\V{E}}{\V{D}}{}}\:\V{X} \mapsto
    \pair{\pair{\DVar{}}{\V{E}}{}}
         {\pair{\DSigma{\V{X}}{\DUnit}}{\V{D}}{}}{}
}\]
%
We simply add a constructor, $\SYMBDVar$, making its arguments
$\DSigma{\V{X}}{\DUnit}$---just an element of $\V{X}$. 
$\V{E}$ and $\V{D}$ stay put, leaving the other constructors
unchanged. Unfolding the interpretation of this definition, we
find an extended sum, corresponding to the \(\V{X}\D{+}\) in
$\FMFreeMonadD$. Taking the fixpoint ties the knot and we have our
data.

@@

@@

Now we need the operations. As expected, \(\LAM{\x}\DVar{\x}\)
plays the r\^ole of \return, making variables terms. Meanwhile,
\bind is indeed \emph{substitution}, which we now
implement generically, making use of $\cata$. Let us write the type,
and start filling in the blanks:
%
\[\stk{
\begin{array}{@{}ll}
\subst : & \PITEL{\V{D}}{\TagDesc}
           \PI{\V{X}, \V{Y}}{\Set} 
           (\V{X} \To \Mu{(\toDesc{(\FreeMonad{\V{D}}{\V{Y}})})}) \To \\
         & \Mu{(\toDesc{(\FreeMonad{\V{D}}{\V{X}})})} \To
           \Mu{(\toDesc{(\FreeMonad{\V{D}}{\V{Y}})})} 
\end{array} \\
\subst\: \V{D}\: \V{X}\: \V{Y}\: \V{\sigma} \mapsto
  \cata\: (\toDesc{(\FreeMonad{\V{D}}{\V{X}})})\: 
          \Mu{(\toDesc{(\FreeMonad{\V{D}}{\V{Y}})})}\: 
          \SHED
}\]
%
We are left with implementing the algebra of the catamorphism. Its
role is to catch appearances of $\DVar{\V{x}}$ and replace them by
$\V{\sigma}\: \V{x}$. This corresponds to the following definition:
%
\[\stk{
\begin{array}{@{}ll}
\apply : & \PITEL{\V{D}}{\TagDesc} 
           \PI{\V{X}, \V{Y}}{\Set} 
           (\V{X} \To \Mu{(\toDesc{(\FreeMonad{\V{D}}{\V{Y}})})}) \To \\
         & \descop{\toDesc{(\FreeMonad{\V{D}}{\V{X}})}}{\Mu{(\toDesc{(\FreeMonad{\V{D}}{\V{Y}})})}} \To
           \Mu{(\toDesc{(\FreeMonad{\V{D}}{\V{Y}})})}
\\
\end{array} \\
\begin{array}{@{}l@{\:\mapsto\:\:}l}
\apply\: \V{D}\: \V{X}\: \V{Y}\: \V{\sigma}\: \pair{\SYMBDVar}{\V{x}}{}   & \V{\sigma}\: \V{x}                   \\
\apply\: \V{D}\: \V{X}\: \V{Y}\: \V{\sigma}\: \pair{\V{c}}{\V{xs}}{} & \Con{\pair{\V{c}}{\V{xs}}{}}
\end{array}
}\]

@@

We complete the hole with $\apply\: \V{D}\: \V{X}\: \V{Y}\:
\V{\sigma}$. Every tagged
description can be seen as a signature of operations: we can uniformly
add a notion of variable, building a new type from an old one, then
providing the substitution structure.
